---
title: "Education Levels and Voter Turnout"
author: "Clare Bodine"
description: "PS 270 Final Project"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction

The research question I will be examining is: **“How do education levels affect voting participation?”** Studying the relationship between education and voting is incredibly important because it helps us to understand political participation and whose opinions are truly reflected in democracy. Education often provides students with the skills and knowledge needed to register and participate, so this analysis could help determine if the voting process is truly representative of our entire population. Significant discrepancies across education levels could insinuate inequality and misrepresentation.

My **hypothesis** is that **higher education levels are positively correlated with voting participation.** This is likely a result of greater access to civic skills and more political knowledge within the education system.


```{r echo=FALSE}
library(tidyverse)
edu_voting <- read_csv("~/Downloads/anes_timeseries_2024_csv_20250808/fpdata.csv")
```

## Data

My analysis utilizes the dataset: **American National Election Studies (ANES) 2024 Report**, which collected data from respondents through surveys before and after the presidential election. I filtered the data to focus specifically on education demographics and voting behavior. After assigning the raw numbers to my specific variables and removing invalid responses, the final dataset consisted of 4930 responses, providing a credible sample.

The **independent variable**, **Education Level**, measures the degree of education a respondent has completed. This is treated as an ordinal variable and is categorized into five distinct levels: (1) Less than high school, (2) High school credential, (3) Some college, no bachelor's, (4) Bachelor's degree, and (5) Graduate degree. 

The **dependent/outcome variable**, **Voting Participation**, measures whether or not a respondent voted in the 2024 presidential election. This is treated as a binary variable, where (1) represents a respondent who voted and (0) represents a respondent who did not vote. The bar plot below shows the distribution of this outcome variable across the 4930 respondents.


```{r echo=FALSE}
edu_voting <- edu_voting |>
  select(education = V241465x, voted = V242095x) |>
  mutate(
    education = ifelse(education < 0, NA, education), 
         voted = ifelse(voted < 0, NA, voted),
         education = factor(education,
                            levels = 1:5, 
                            labels = c(
                              "Less than high school",
                              "High school credential", 
                              "Some college, no bachelor's",
                              "Bachelor's degree",
                              "Graduate degree"))
    ) |>
  drop_na()
```


```{r echo = FALSE}
turnout_graph <- edu_voting
ggplot(edu_voting, aes(x = factor(voted, labels = c("Did Not Vote", "Voted")))) +
  geom_bar(fill = "salmon") +
  labs(
    title = "Distribution of Voting Participation",
    x = "Voting Behavior",
    y = "Number of Respondents"
  ) +
  theme_minimal()
turnout_graph
```
To determine criteria for supporting or rejecting my hypothesis, I will compare voter turnout percentages within each education level. Data that shows an increase in turnout percentage in line with an increase in education level will support my hypothesis. Data that is inconsistent across each education level without a distinct pattern, or voter turnout percentages that decrease with higher education level will reject my hypothesis. 



## Results 

To interpret the voting participation variable, or voter turnout, I determined the percent of respondents that voted within each distinct education level. The table below expresses the specific turnout rates, showing a clear increase in each level of education, starting from 53.8% for "Less than a high school" to 92.9% for "Graduate degree."


```{r echo=FALSE}
voting_summary <- edu_voting |>
  group_by(education) |>
  summarize(
    Total = n(),
    Voted = sum(voted == 1),
    Not_Voted = sum(voted == 0),
    Turnout_Pct = round(Voted / Total * 100, 1)
  )

voting_summary |>
  knitr::kable(col.names = c ("Education", "Total", "Voted", "Didn't Vote", "Turnout Percent"))

```


This **bar plot** visualizing the relationship between education level and voter turnout percentages further demonstrates that increase. The visual shows a consistent upward trend in voter percentage as education level increases.


```{r echo=FALSE}
result_graph <- voting_summary |>
  ggplot(mapping = aes(x = education, 
                       y = Turnout_Pct,
                       fill = education)) + 
  geom_col() +
  labs(
    title = "Education Levels and Voter Turnout",
    x = "Education Level",
    y = "Percent that Voted"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
                                   legend.position = "none")
result_graph
```
### Linear Regression Model:

```{r echo=FALSE}
edu.reg <- lm(voted ~ as.numeric(education), data = edu_voting)
summary(edu.reg)
```

```{r echo=FALSE}
library(broom)
tidy(edu.reg) |>
  knitr::kable(digits = 3)
```

```{r echo=FALSE}
var_labels <- c("as.numeric(education)" = "Education Effect")
modelsummary::modelsummary(
  list(edu.reg),
  coef_map = var_labels,
  gof_mop = c("nobs", "r.squared", "adj.r.squared"))
```
**Interpretation:**

The results of my linear regression model indicate a statistically significant correlation between education level and voter turnout. This is confirmed by the **p-value of < 2.2e-16 (recorded as 0)**, which means that the likelihood this correlation happened by chance is so small it is nearly zero. 
My main coefficient of interest is the **Education Effect** which is **0.085**. This means that for every increase in one level of education, the probability of a respondent voting increases by 8.5 percentage points. With an **R-squared of 0.066**, we can determine that education level explains 6.6% of voting behavior in this sample.
Although this interpretation provides highly statistically significant results, we should **not interpret them causally** as there are other factors that must be taken into account.


## Conclusion
Overall, my results strongly support my hypothesis that higher education levels are positively correlated with voting participation. The data exemplifies an increase in voter turnout as education levels rise, and the regression model confirms a statistically significant increase of 8.5 percentage points for each education level reached. However, there are limitations to this analysis including potential confounding variables such as income, upbringing, overall political interest, etc. These confounders that effect both education level and voting behavior are threats to inference, as well as my removal of voting variable responses that did not fit "Vote" or "Did not Vote." With more time and resources, I would improve this analysis by incorporating additional variables that may impact voting behavior, creating a more detailed and precise analysis of what factors effect voter turnout in 2024. 


## Code
```{r print-code, ref.label=knitr::all_labels(), echo = TRUE, eval=FALSE}

```

### AI Disclosure
I utilized AI throughout my final project to assist with difficult coding chunks and troubleshooting. I used AI in my initial data cleaning, asking specific questions to turn the raw data into variables and condense the dataset. From there, AI served as a tool for fixing errors in my code for visuals and interpretation of results when I was stuck.